---
title: "Notebook for tidying up the data file from the Common Garden counting season 2021"
author: "Dominik Ziaja"
output:
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    number_section: true
    df_print: paged
    theme: cerulean
    
---

# Load in

## Packages
```{r}
library(naniar)
library(dplyr)
library(tidyr)
library(stringr)
```
## Data
```{r}
df <- read.csv("Data/Counteddata_ComGar_2021.csv",
               sep = ",",
              na = c("", "NA"),
              header=T)
```

# Checking of dataframe

**Get the summary to see whether something is odd**
```{r}
summary(df)
```

**Visualise NA values**

```{r}
vis_miss(df)
```

No na.values observed where no na values were expected. Since during the read in empty cells were marked as
"NA", the NA values of the aphid columns are simply zeros which will later be transformed into it.
Important is, that no NA values are in the Metainformation-columns, which there are not.

**check whether there are wrong/cryptic names in the meta columns**
```{r}
for (i in names(df)){
  cat(c("The unique values of: ", i, "are the following:"))
  print(unique(df[[i]]))
  cat("\n")
  }

```
All unique values in the columns make sense. Quite a lot of NAs probably due to
a lot of 0 values not typed in. Often the last line of a read in file
also contains NA values

# Data wrangling/tidying

## Transform Plot and Position into unique coding
It allows to directly observe partially crossed/nested data and also allows us
to calculate the nestedness by the glmm without specifically using the
syntax for it. Therefore minimizing the possible errors made during the
calculations

Instead of reusing Plot 1-10 6 times, they will be numbered 1-60,
the same applies to the positions. Instead of 1-5 x 60, we will use the
numbers 1-300 to really show that all positions are unique.

**Set up the vectors to insert unique positions**

```{r}
# setting up the new positions
vector <- seq(1,300,1)
Positions <- c()

for (i in 1:length(unique(df$Week))){
  Positions <- c(Positions, vector)
}
```



```{r}
df.tidied <- df %>%
  mutate_all(~replace(., is.na(.), 0)) %>% # set all na values to 0
  mutate(Week = Week + 17) %>% # add 17 to the calendar weeks in order to fit the calendar week (start at 18 instead of 1)
  mutate(Plot = case_when(Block == 1 ~ Plot, # make the plot have unique coding, right now it's redundant
                          Block == 2 ~ Plot + 10,
                          Block == 3 ~ Plot + 20,
                          Block == 4 ~ Plot + 30,
                          Block == 5 ~ Plot + 40,
                          Block == 6 ~ Plot + 50,)) %>%
  mutate(Position = Positions) %>% # Only works if dataframe is sorted after Block-Plot-Position. Very prone to errors
  filter(Plot != 1) %>% # remove here already Plot 1 since one plant of a wrong chemotype was planted into this plot
  mutate(Plottype = case_when(Plottype == 0 ~ "Homogenous",
                              Plottype == 1 ~ "Heterogenous"))
```


**Put a column called "Plant_standardized" into the dataframe to match with other files late ron**

```{r}
df.tidied <- df.tidied %>%
separate("Plant", # make multiple columns out of the Plant column
           into = c("Motherplant_ID", "Plant_number", "Plottype_abbreviation"),
           sep = "_",
         remove = FALSE) %>%
  mutate(Motherplant_ID = str_pad(Motherplant_ID, width = 2, side = "left", pad = "0"),  # pad Motherplant with zeros to width = 2
         Plant_number = str_pad(Plant_number, width = 3, side = "left", pad = "0")) %>%  # pad the plant number with zeros to width = 3
  unite(., Plant_standardized, c("Motherplant_ID", "Plant_number", "Plottype_abbreviation"), sep = "_") # unite again to one column
```

## Checking the data types of the different columns
Sometimes R reads in integer values as characters instead of values
```{r}
for (i in 1:ncol(df.tidied)) {
  print(paste("The charactertype of ", colnames(df.tidied)[i], "is: ", typeof(df.tidied[, i])))
}
```
Only the columns "Ant_total" and "U_tanaceti_apterous" need to be changed to integer.
Also, the column "Date" needs to be formatted as date.

However, the coercion resulted in introduction of NA since some values are not
1:1 convertable. Especially the count of ants close to a nest were defined as
">50" on 4 occasions. Therefore, these will be in this case replaced to the
value 50 which will from now on be defined as the maximum measurement.
The same as the value 2000 of the apterous uroleucon count.

## Set Ant values and reformat certain columns
```{r}
df.tidied <- df.tidied %>%
  mutate(Ant_total = ifelse(Ant_total == ">50", 50, Ant_total),
         U_tanaceti_apterous = as.integer(U_tanaceti_apterous),
         Ant_total = as.integer(Ant_total),
         Date = as.Date(Date, format = "%d.%m.%Y"))
```

## Adding factor levels to the chemotypes to define the order for ggplotting
The defined factor levels follow the magnitude of average alpha-Diversity
of the different chemotypes. From the lowest (Keto) to the highest (Myrox)

```{r}

df.tidied$Chemotype = factor(df.tidied$Chemotype, levels=c("Keto",
                                             "BThu",
                                             "ABThu",
                                             "Aacet",
                                             "Myrox"))

```

## Add a column containing the motherplant information
```{r}
df.tidied <- df.tidied %>%
  mutate(Motherplant = as.character(stringr::str_split_fixed(Plant, "_", 2)[,1]))

```


All values were 0 before the checked dates. So no values need to be replaced.
However, the colonisation events of Uroleucon_tanaceti need to be set to 0, if
winged nymphs were observed. Because then, it is very unlikely that the plant was still colonised
by winged aphids not emerged from the plant.

## Adding columns for total counts of taxa

```{r}
df.tidied <- df.tidied %>%
  mutate(M_fuscoviride_total = M_fuscoviride_alate + M_fuscoviride_apterous,
         M_fuscoviride_total_presence = ifelse(M_fuscoviride_total > 0, 1, 0),
         U_tanaceti_total = U_tanaceti_alate + U_tanaceti_apterous,
         U_tanaceti_total_presence = ifelse(U_tanaceti_total > 0, 1, 0),
         M_tanacetaria_total = M_tanacetaria_alate + M_tanacetaria_apterous,
         M_tanacetaria_total_presence = ifelse(M_tanacetaria_total > 0, 1, 0),
         Ant_total_presence = ifelse(Ant_total > 0, 1, 0),
         Ant_MF_presence = ifelse(Ant_MF > 0, 1, 0),
         Antnest_presence = ifelse(Antnest == 0, 0, 1))


```


## Setting all Uroleucon alate values to 0 when Uroleucon alate nymphs were present
This is specifically done after the total value columns were calculated to 
incorporate the count of alate into the statistical analysis regarding the
total count.

```{r}
df.tidied <- df.tidied %>%
  mutate(U_tanaceti_alate = ifelse(U_tanaceti_WN == 1, 0, U_tanaceti_alate))
```


## Adding columns calculating the alate presence
Adding the alate presence specifically after the alate count was set to 0 if winged
nymphs were observed. This was done since this metric is a proxy for the colonisation of plants
by alate.

```{r}
df.tidied <- df.tidied %>%
  mutate(
    M_fuscoviride_alate_presence = ifelse(M_fuscoviride_alate > 0, 1, 0),
    U_tanaceti_alate_presence = ifelse(U_tanaceti_alate > 0, 1, 0),
    M_tanacetaria_alate_presence = ifelse(M_tanacetaria_alate > 0, 1, 0)
  )
```


## Reordering all the columns into a correct order
```{r}
df.tidied <- df.tidied %>%
  relocate("Motherplant", .after = "Plant_standardized") %>%
  relocate(contains("M_fuscoviride"), .after = "Plottype") %>%
  relocate(contains("M_tanacetaria"), .after = "M_fuscoviride_total_presence") %>%
  relocate(contains("U_tanaceti"), .after = "M_tanacetaria_total_presence") %>%
  relocate(contains("Ant"), .after = "U_tanaceti_total_presence") %>%
  relocate(c("Plant", "Plant_standardized", "Motherplant"), .after = "Position") %>%
  select(-c("Notes", "Ant_MF", "Antnest", "Ant_total", "U_tanaceti_WN"))
```


## Saving the dataframe into the directory
```{r}
write.csv(df.tidied, "Data/Counteddata_ComGar_2021_tidied.csv", row.names = F)
```
