---
title: "Descriptive Plots Common Garden 2021 aphid counting"
author: "Dominik Ziaja"
output:
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    number_section: true
    df_print: paged
    theme: cerulean
---

# Preparation
```{r}
library(naniar)
library(dplyr)
library(ggplot2)
library(tidyr)
library(patchwork)
```
# Vectors used throughout the script

```{r}
colours_chemotype <-  c("#FFC466", "#77A8F5", "#F9623E", "#6ECFBD", "#B481FF")
labels_chemotype <- c("Keto", "BThu", "ABThu", "Aacet", "Myrox")
```


# Reading in and checking Data

```{r}
df <- read.csv("Data/Counteddata_ComGar_2021_tidied.csv", header=T)
```

**Visualise NA values**

```{r}
vis_miss(df)
```

No NA values present in the dataframe.

**Check in the summary if anything abnormal is there**

```{r}
summary(df)
```

# Data wrangling

## Adding factor levels 

**Set levels of the plot-type for plotting later on**
```{r}
df$Plottype <- factor(df$Plottype, levels = c("Homogenous", "Heterogenous"))
```

The defined factor levels follow the magnitude of average alpha-Diversity
of the different chemotypes. From the lowest (Keto) to the highest (Myrox)

```{r}
df$Chemotype = factor(df$Chemotype, levels=c("Keto",
                                             "BThu",
                                             "ABThu",
                                             "Aacet",
                                             "Myrox"))
```


# Summarising sum and mean of dataframe
First we need to summarize on the treatmentlevel (Chemotype, Plottype) per week

```{r}
df.sum <- df %>%
  group_by(Chemotype, Plottype, Week) %>%
  summarise(across(where(is.numeric), ~ sum(.x, na.rm = TRUE)))
```


```{r}
df.mean <- df %>%
  group_by(Chemotype, Plottype, Week) %>%
  summarise(across(where(is.numeric), ~ mean(.x, na.rm = TRUE)))
```

# Plotting the different descriptive plots

## Presence of all aphid species together
First setting up the dataframe containing only the info we need/want
```{r}
df.sum.presence <- df.sum %>%
  select(c("Chemotype":"Motherplant",
           "U_tanaceti_total_presence",
           "M_fuscoviride_total_presence",
           "M_tanacetaria_total_presence")) %>%
  pivot_longer(.,
               cols = U_tanaceti_total_presence:M_tanacetaria_total_presence,
               names_to = "metric",
               values_to = "count")

df.sum.presence$metric <- factor(df.sum.presence$metric, 
                                 levels = c("U_tanaceti_total_presence",
                                            "M_tanacetaria_total_presence",
                                            "M_fuscoviride_total_presence"))
```

Then plotting the actual plot faceted for Plot-type
```{r}
gg.presence.allspecies.sum <- ggplot(data=df.sum.presence, aes(x=Week, y=count, color=Chemotype,
                                               shape=metric, linetype=metric)) +
  geom_line(size=.8) +
  geom_point(size=1.7) +
  scale_colour_manual(values = colours_chemotype,
                      labels = labels_chemotype) +
  facet_grid(Chemotype ~ Plottype) +
  theme_light(base_size=9) +
  labs(color = "Chemotype", linetype="", shape="") +
    scale_x_continuous("\n\nCalendar week 2021",
                     limits=c(min(df.sum.presence$Week), max(df.sum.presence$Week)),
                     breaks = seq(min(df.sum.presence$Week), max(df.sum.presence$Week), 2)) +
  ylab("Abundance\n") +
  scale_linetype_manual(values = c(1,2,3),
                        labels = c("U. tanaceti",
                                   "M. tanacetaria",
                                   "M. fuscoviride")) +
  scale_shape_manual(values = c(15, 16, 17),
                     labels = c("U. tanaceti",
                                "M. tanacetaria",
                                "M. fuscoviride")) +
    theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        text = element_text(size=9),
        strip.text.y = element_blank(),
        strip.text.x = element_text(
          size = 9, color = "black"),
        strip.background = element_rect(
          color = "white", fill = "white"),
        legend.text = element_text(size=10),
        legend.title = element_text(size=11),
        legend.key.height = unit(1, "cm"),
        legend.key.width = unit(1.8,"cm")) +
  annotate("rect", xmin = 18, xmax = 21,
           ymin = 0, ymax = 30,
           alpha = .3, fill = "grey") +
  annotate("rect", xmin = 26, xmax = 30,
           ymin = 0, ymax = 30,
           alpha = .3, fill = "grey") +
  annotate(geom = "text",
           x = 19.5,
           y = 0,
           vjust = 3.6,
           label = "May",
           size = 3) +
  annotate(geom = "text",
           x = 23.5,
           y = 0,
           vjust = 3.6,
           label = "Jun",
           size = 3) +
  annotate(geom = "text",
           x = 28,
           y = 0,
           vjust = 3.6,
           label = "Jul",
           size = 3) +
  annotate(geom = "text",
           x = 32,
           y = 0,
           vjust = 3.6,
           label = "Aug",
           size = 3) +
  coord_cartesian(xlim = c(18, 33), clip = "off")

plot(gg.presence.allspecies.sum)

ggsave(filename = "Plots/Figure_S2.pdf", 
       plot = gg.presence.allspecies.sum,
       dpi = 600,
       device = "pdf",
       width = 18,
       height = 14,
       units = "cm")

ggsave(filename="Plots/Figure_S2.jpg",
       plot = gg.presence.allspecies.sum,
       dpi = 600,
       width = 18,
       height = 14,
       units = "cm")

```


## Mean total count *U. tanaceti*

First, set up the dataframe with the necessary metrics

```{r}
df.mean.UT <- df.mean %>%
  select(c("Chemotype":"Motherplant",
           "U_tanaceti_total")) %>%
  pivot_longer(.,
               cols = U_tanaceti_total,
               names_to = "metric",
               values_to = "count")

```

Then plot the plot before later merging it
```{r fig.height=10, fig.width=10}
gg.total.ut.mean <- ggplot(data=df.mean.UT, aes(x=Week, y=count, color=Chemotype,
                                shape=Plottype, linetype = Plottype)) +
  geom_line(size=0.8) +
  geom_point(size=1.5) +
  scale_colour_manual(values = colours_chemotype,
                      labels = labels_chemotype) +
  theme_light() +
  labs(color = "Chemotype", linetype="", shape="") +
    scale_x_continuous(limits=c(min(df.mean.UT$Week), max(df.mean.UT$Week)),
                     breaks = seq(min(df.mean.UT$Week), max(df.mean.UT$Week), 2)) +
  facet_wrap(~ Chemotype, nrow=5) +
  scale_shape_manual(values = c(15, 16),
                     name="Plot-type") +
  scale_linetype_manual(values=c("solid", "dashed"),
                        name="Plot-type") +
  ylab("Abundance\n") +
  xlab("") +
    guides(color = guide_legend(order=1),
         shape = guide_legend(order=2),
         linetype = guide_legend(order=2)) +
    theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        text = element_text(size=10),
        strip.text = element_blank(),
        legend.text = element_text(size=8),
        legend.title = element_text(size=8),
        axis.title.y = element_text(size = 11),
        legend.key.height = unit(1, "cm"),
        legend.key.width = unit(1,"cm")) +
  geom_vline(xintercept = 25, color="red4",
             alpha = 0.3) +
  annotate("rect", xmin = 18, xmax = 21,
           ymin = 0, ymax = 630,
           alpha = .3, fill = "grey") +
  annotate("rect", xmin = 26, xmax = 30,
           ymin = 0, ymax = 630,
           alpha = .3, fill = "grey") +
  annotate(geom = "text",
           x = 19.5,
           y = 0,
           vjust = 4,
           label = "May",
           size = 3) +
  annotate(geom = "text",
           x = 23.5,
           y = 0,
           vjust = 4,
           label = "Jun",
           size = 3) +
  annotate(geom = "text",
           x = 28,
           y = 0,
           vjust = 4,
           label = "Jul",
           size = 3) +
  annotate(geom = "text",
           x = 32,
           y = 0,
           vjust = 4,
           label = "Aug",
           size = 3) +
  coord_cartesian(xlim = c(18, 33), clip = "off")


show(gg.total.ut.mean)
```

## Mean total count *M. tanacetaria*

First, set up the dataframe with the necessary metrics

```{r}
df.mean.MT <- df.mean %>%
  select(c("Chemotype":"Motherplant",
           "M_tanacetaria_total")) %>%
  pivot_longer(.,
               cols = M_tanacetaria_total,
               names_to = "metric",
               values_to = "count")

```


```{r fig.height=10, fig.width=10}

gg.total.mt.mean <- ggplot(data=df.mean.MT, aes(x=Week, y=count, color=Chemotype,
                                shape=Plottype, linetype = Plottype)) +
  geom_line(size=0.8) +
  geom_point(size=1.5) +
  scale_colour_manual(values = colours_chemotype,
                      labels = labels_chemotype) +
  theme_light() +
  labs(color = "Chemotype", linetype="", shape="") +
    scale_x_continuous(
                     limits=c(min(df.mean.MT$Week), max(df.mean.MT$Week)),
                     breaks = seq(min(df.mean.MT$Week), max(df.mean.MT$Week), 2)) +
  facet_wrap(~ Chemotype, nrow=5) +
  scale_shape_manual(values = c(15, 16),
                     name="Plot-type") +
  scale_linetype_manual(values=c("solid", "dashed"),
                        name="Plot-type") +
  ylab("Abundance\n") +
    theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        text = element_text(size=10),
        strip.text = element_blank(),
        axis.title.x = element_text(size = 11),
        axis.title.y = element_text(size = 11)) +
     xlab("\nCalendar week 2021") +
     ylab("") +
    guides(color = guide_legend(order=1),
         shape = guide_legend(order=2),
         linetype = guide_legend(order=2)) +
  geom_vline(xintercept = 25, color="green4",
             alpha=0.3) +
  annotate("rect", xmin = 18, xmax = 21,
           ymin = 0, ymax = 45,
           alpha = .3, fill = "grey") +
  annotate("rect", xmin = 26, xmax = 30,
           ymin = 0, ymax = 45,
           alpha = .3, fill = "grey") +
  annotate(geom = "text",
           x = 19.5,
           y = 0,
           vjust = 4,
           label = "May",
           size = 3) +
  annotate(geom = "text",
           x = 23.5,
           y = 0,
           vjust = 4,
           label = "Jun",
           size = 3) +
  annotate(geom = "text",
           x = 28,
           y = 0,
           vjust = 4,
           label = "Jul",
           size = 3) +
  annotate(geom = "text",
           x = 32,
           y = 0,
           vjust = 4,
           label = "Aug",
           size = 3) +
  coord_cartesian(xlim = c(18, 33), clip = "off")

show(gg.total.mt.mean)
```

## Mean total count *M. fuscoviride*

First, set up the dataframe with the necessary metrics

```{r}
df.mean.MF <- df.mean %>%
  select(c("Chemotype":"Motherplant",
           "M_fuscoviride_total")) %>%
  pivot_longer(.,
               cols = M_fuscoviride_total,
               names_to = "metric",
               values_to = "count")

```


```{r fig.height=10, fig.width=10}

gg.total.mf.mean <- ggplot(data=df.mean.MF, aes(x=Week, y=count, color=Chemotype,
                                shape=Plottype, linetype = Plottype)) +
  geom_line(size=0.8) +
  geom_point(size=1.5) +
  scale_colour_manual(values = colours_chemotype,
                      labels = labels_chemotype) +
  theme_light() +
  labs(color = "Chemotype", linetype="", shape="") +
    scale_x_continuous(#"\n\nCalendar week 2021",
                     limits=c(min(df.mean.MF$Week), max(df.mean.MF$Week)),
                     breaks = seq(min(df.mean.MF$Week), max(df.mean.MF$Week), 2)) +
  facet_wrap(~ Chemotype, nrow=5) +
  scale_shape_manual(values = c(15, 16),
                     name="Plot-type") +
  scale_linetype_manual(values=c("solid", "dashed"),
                        name="Plot-type") +
  ylim(0,65) +
    theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        text = element_text(size=10),
        strip.text = element_blank()) +
    xlab("") +
    ylab("") +
    guides(color = guide_legend(order=1),
         shape = guide_legend(order=2),
         linetype = guide_legend(order=2)) +
  geom_vline(xintercept = 25, color="hotpink4",
             alpha=0.3) +
  annotate("rect", xmin = 18, xmax = 21,
           ymin = 0, ymax = 65,
           alpha = .3, fill = "grey") +
  annotate("rect", xmin = 26, xmax = 30,
           ymin = 0, ymax = 65,
           alpha = .3, fill = "grey") +
  annotate(geom = "text",
           x = 19.5,
           y = 0,
           vjust = 4,
           label = "May",
           size = 3) +
  annotate(geom = "text",
           x = 23.5,
           y = 0,
           vjust = 4,
           label = "Jun",
           size = 3) +
  annotate(geom = "text",
           x = 28,
           y = 0,
           vjust = 4,
           label = "Jul",
           size = 3) +
  annotate(geom = "text",
           x = 32,
           y = 0,
           vjust = 4,
           label = "Aug",
           size = 3) +
  coord_cartesian(xlim = c(18, 33), clip = "off")
  
show(gg.total.mf.mean)

```

# Merging the total mean plots of all three species together

```{r fig.width=7.08, fig.height=5.51}

gg.total.ut.mean.merge <- gg.total.ut.mean + theme(legend.key.width = unit(1.4, "cm"),
                                                   legend.key.size = unit(1.2, "cm"),
                                                   text = element_text(size = 9),
                                                   legend.title = element_text(size = 10),
                                                   legend.text = element_text(size = 10))
gg.total.mt.mean.merge <- gg.total.mt.mean + theme(text = element_text(size = 9))
gg.total.mf.mean.merge <- gg.total.mf.mean + theme(text = element_text(size = 9))

gg.all.species.total.mean <-  gg.total.ut.mean.merge +
  gg.total.mt.mean.merge + guides(colour = "none", linetype = "none", shape = "none") +
  gg.total.mf.mean.merge + guides(colour = "none", linetype = "none", shape = "none") & theme(legend.position = "right")
gg.all.species.total.mean <- gg.all.species.total.mean + plot_layout(guides = "collect") & theme(legend.text = element_text(size = 8))

show(gg.all.species.total.mean)

ggsave(filename = "Plots/Figure_1.pdf", 
       plot = gg.all.species.total.mean,
       dpi = 600,
       device = "pdf",
       width = 18,
       height = 14,
       units = "cm")

ggsave(filename = "Plots/Figure_1.jpg", 
       plot = gg.all.species.total.mean,
       dpi = 600,
       width = 18,
       height = 14,
       units = "cm")
```

# Analysis of cumulative sum of the different aphid species
Calculating the percentage wise cumulative sum of the binary metrics
```{r}
df.cumsum <- df.sum %>%
  group_by(Week) %>%
  summarise(across(where(is.numeric), ~ sum(.x, na.rm = TRUE))) %>%
 mutate(
   UT_total = cumsum(U_tanaceti_total),
   MT_total = cumsum(M_tanacetaria_total),
   MF_total = cumsum(M_fuscoviride_total),
   UT_alate_presence = cumsum(U_tanaceti_alate_presence),
   MT_alate_presence = cumsum(M_tanacetaria_alate_presence),
   MF_alate_presence = cumsum(M_fuscoviride_alate_presence),
   UT_alate_count = cumsum(U_tanaceti_alate),
   MT_alate_count = cumsum(M_tanacetaria_alate),
   MF_alate_count = cumsum(M_fuscoviride_alate),
   UT_total_presence = cumsum(U_tanaceti_total_presence),
   MT_total_presence = cumsum(M_tanacetaria_total_presence),
   MF_total_presence = cumsum(M_fuscoviride_total_presence)
   ) %>%
  mutate(
    UT_total = UT_total/max(UT_total)*100,
    MT_total = MT_total/max(MT_total)*100,
    MF_total = MF_total/max(MF_total)*100,
    UT_alate_presence = UT_alate_presence/max(UT_alate_presence)*100,
    MT_alate_presence = MT_alate_presence/max(MT_alate_presence)*100,
    MF_alate_presence = MF_alate_presence/max(MF_alate_presence)*100,
    UT_alate_count = UT_alate_count/max(UT_alate_count)*100,
    MT_alate_count = MT_alate_count/max(MT_alate_count)*100,
    MF_alate_count = MF_alate_count/max(MF_alate_count)*100,
    UT_total_presence = UT_total_presence/max(UT_total_presence)*100,
    MT_total_presence = MT_total_presence/max(MT_total_presence)*100,
    MF_total_presence = MF_total_presence/max(MF_total_presence)*100
    ) %>%
   select(c("Week","UT_total":"MF_total_presence")) %>%
    pivot_longer(.,
               cols = UT_total:MF_total_presence,
               names_to = "metric",
               values_to = "cumulative_sum")
```

## Cumulative sum *U. tanaceti*
```{r}

df.cumsum.UT <- df.cumsum %>%
  filter(grepl("UT", metric)) # subset for only values of UT

gg.cumsum.UT <- ggplot(df.cumsum.UT) +
  geom_line(aes(x=Week, y=cumulative_sum, color=metric)) +
  geom_point(aes(x=Week, y=cumulative_sum, shape=metric), size=2) +
    scale_x_continuous("\nCalendar week 2021",
                     limits=c(min(df.cumsum$Week), max(df.cumsum$Week)),
                     breaks = seq(min(df.cumsum$Week), max(df.cumsum$Week), 1)) +
  scale_color_manual(values=c("black", "black", "black", "black", "black")) +
  ylab("\nCumulative sum [%]") +
  guides(color="none") +
  theme_classic()

show(gg.cumsum.UT)
```

## Cumulative sum *M. tanacetaria* 

```{r}

df.cumsum.MT <- df.cumsum %>%
  filter(grepl("MT", metric)) # subset for only values of MT

gg.cumsum.MT <- ggplot(df.cumsum.MT) +
  geom_line(aes(x=Week, y=cumulative_sum, color=metric)) +
  geom_point(aes(x=Week, y=cumulative_sum, shape=metric), size=2) +
    scale_x_continuous("\nCalendar week 2021",
                     limits=c(min(df.cumsum$Week), max(df.cumsum$Week)),
                     breaks = seq(min(df.cumsum$Week), max(df.cumsum$Week), 1)) +
  scale_color_manual(values=c("black", "black", "black", "black", "black", "black")) +
  ylab("\nCumulative sum [%]") +
  guides(color="none") +
  theme_classic()

show(gg.cumsum.MT)
```

## cumulative sum *M. fuscoviride*

```{r}

df.cumsum.MF <- df.cumsum %>%
  filter(grepl("MF", metric)) # subset for only values of MF

gg.cumsum.MF <- ggplot(df.cumsum.MF) +
  geom_line(aes(x=Week, y=cumulative_sum, color=metric)) +
  geom_point(aes(x=Week, y=cumulative_sum, shape=metric), size=2) +
    scale_x_continuous("\nCalendar week 2021",
                     limits=c(min(df.cumsum$Week), max(df.cumsum$Week)),
                     breaks = seq(min(df.cumsum$Week), max(df.cumsum$Week), 1)) +
  scale_color_manual(values=c("black", "black", "black", "black", "black")) +
  ylab("\nCumulative sum [%]") +
  guides(color="none") +
  theme_classic()

show(gg.cumsum.MF)
```