---
title: "Script calculating the GLMMs for the aphid data spanning the whole season 2021"
author: "Dominik Ziaja"
output:
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    number_section: true
    df_print: paged
    theme: cerulean
---

# Load in necessities

## Packages used
```{r}
library("glmmTMB")
library("ggplot2")
library("dplyr")
library("DHARMa")
library("car")
library("insight")
library("emmeans")
library("patchwork")
## cosmetic
theme_set(theme_bw()+
theme(panel.spacing=grid::unit(0,"lines")))
```

## Setting up colors for the chemotypes used throughout the script

```{r}
colours_chemotype <-  c("#FFC466", "#77A8F5", "#F9623E", "#6ECFBD", "#B481FF")
labels_chemotype <- c("Keto", "BThu", "ABThu", "Aacet", "Myrox")
```


## Reading in the data ## 
```{r}
data <- read.csv("Data/Counteddata_ComGar_2021_tidied.csv")
```

# Data wrangling

## Set up a column with the Clone_ID
Using clone ID as a random factor in order to account for their variance since
otherwise differences based on Plottype might be caused false-positive
```{r}
data <- data %>%
  mutate(Clone_ID = as.character(
    paste(
    stringr::str_split_fixed(data$Plant, "_", 3)[,1], # Motherplant ID
    "_", # Underscore to combine
    stringr::str_split_fixed(data$Plant, "_", 3)[,2], # F1-ID
    sep="")))
```

A small check, whether what we wanted to do actually worked.
Testing on 20 random plant labels in the dataset
```{r}
for (i in sample.int(length(unique(data$Plant)), 20)){
  print(cat(paste("Plant_ID: ", data$Plant[i], " Clone_ID: ", data$Clone_ID[i]), "/n "))}
```
# *U. tanaceti*

## Total count

Filtering of weeks based on the cumulative plot for UT total count
```{r}
data.total.UT <- data %>%
  filter(Week > 20,
         Week < 29)
```

### The zero inflation model
```{r}
UT.total.glmm.zi.complex <- glmmTMB(U_tanaceti_total ~ Chemotype * Plottype + Week
                    + (1|Motherplant) 
                    + (1|Clone_ID) 
                    + (1|Block) 
                    + (1|Plot) 
                    + (1|Counted_by) 
                    + (1|Plant),
                    family = "poisson",
                    zi=~ Chemotype * Plottype + Week
                    + (1|Motherplant) 
                    + (1|Clone_ID) 
                    + (1|Block) 
                    + (1|Plot) 
                    + (1|Counted_by) 
                    + (1|Plant),
                    data=data.total.UT)
```


### Diagnostics
**Simulate and plot residuals**
```{r}
sim.UT.total.glmm.zi.complex <- simulateResiduals(fittedModel = UT.total.glmm.zi.complex, n=1000)
plot(sim.UT.total.glmm.zi.complex, quantreg=T)
```



### Significances

**Receive the variance components of the model**
```{r}
var.glmm.ut.zi.complex <- get_variance(UT.total.glmm.zi.complex)
var.glmm.ut.zi.complex
```

**Chi-square test**
```{r}
anova.var.glmm.zi.complex <- Anova(UT.total.glmm.zi.complex, type="III",
                                          test="Chisq")

anova.var.glmm.zi.complex.zi <- glmmTMB:::Anova.III.glmmTMB(UT.total.glmm.zi.complex,
                                                            component = "zi")

print(anova.var.glmm.zi.complex, digits = 3)
print(anova.var.glmm.zi.complex.zi, digits = 3)
```

**Get the emmeans estimates**

```{r}
UT.count.complex.emmeans <- as.data.frame(emmeans(UT.total.glmm.zi.complex, pairwise~Chemotype*Plottype, type="response")$emmeans)

UT.count.complex.emmeans$Chemotype <- factor(UT.count.complex.emmeans$Chemotype,
                                             levels = c("Keto",
                                                        "BThu",
                                                        "ABThu",
                                                        "Aacet",
                                                        "Myrox"))
UT.count.complex.emmeans$Plottype <- factor(UT.count.complex.emmeans$Plottype,
                                            levels = c("Homogenous", "Heterogenous"))

UT.count.complex.emmeans <- UT.count.complex.emmeans %>%
  mutate(errorbarmin = rate  - SE,
         errorbarmax = rate + SE )
```

**Plot the emmeans estimates**

```{r}
barplot.lsmeans.ut.total <- ggplot(aes(x=Chemotype,
                                       y=rate,
                                       fill=Chemotype,
                                       alpha=Plottype),
                                   data= UT.count.complex.emmeans) +
  geom_bar(stat="identity", position="dodge") +
  geom_errorbar(aes(ymin = rate-SE,
                    ymax=rate+SE),
                width=.3,
                position=position_dodge(.9)) +
  ylim(0,6) +
  scale_x_discrete("\nChemotype") +
  scale_fill_manual(values = colours_chemotype, guide = "none") +
  labs(alpha="Plot-type") +
  scale_y_continuous(expression(paste("Total count  ", italic("U. tanaceti")))) +
  guides(alpha = guide_legend(override.aes = list(fill="grey",
                                                  alpha = c(1, 0.5),
                                                  linetype = 0))) +
  scale_alpha_discrete(range = c(1,0.4)) +
  theme_classic() +
  theme(text = element_text(size=14),
        axis.title.y = element_text(vjust = +3))

show(barplot.lsmeans.ut.total)

```

## Alate

Filtering the dataset for certain weeks based on the cumulative plot
```{r}
df.alate.UT <- data %>%
  dplyr::filter(Week > 21,
         Week < 30)
```


### Binomial model

```{r}
glmm.UT.alate.presence <- glmmTMB(U_tanaceti_alate_presence ~ Chemotype*Plottype
                                  + Week 
                                  + (1|Block) 
                                  + (1|Plot) 
                                  + (1|Clone_ID) 
                                  + (1|Motherplant) 
                                  + (1|Counted_by) 
                                  + (1|Plant),
                    data=df.alate.UT,
                    family="binomial")
```


### Diagnostics

DHARMA residual plots
```{r}
sim.UT.alate.presence <- simulateResiduals(fittedModel = glmm.UT.alate.presence, n=1000)
plot(sim.UT.alate.presence, quantreg=T)
```

### Significances

**receive the variance of each factor of the model**
```{r}
var.glmm.binomial.alate.presence <- get_variance(glmm.UT.alate.presence, tolerance = 1e-11)

print(var.glmm.binomial.alate.presence)

```

**Chi-square test**
```{r}
anova.glmm.binomial.alate.presence <- Anova(glmm.UT.alate.presence, type="III",
                                          test="Chisq")

print(anova.glmm.binomial.alate.presence, digits = 3)
```


# *Macrosiphoniella tanacetaria* #


## Total count
```{r}
df.MT.total <- data %>%
  filter(Week > 18,
         Week < 27)
```

### The zero-inflation model
```{r}
MT.total.glmm.zi.complex <- glmmTMB(M_tanacetaria_total ~ Chemotype * Plottype + Week
                    + (1|Block)
                    + (1|Plot)
                    + (1|Clone_ID)
                    + (1|Counted_by)
                    + (1|Plant),
                    family = "poisson",
                    zi=~Chemotype * Plottype + Week 
                    + (1|Block) 
                    + (1|Plot)  
                    + (1|Clone_ID)
                    + (1|Counted_by)
                    + (1|Plant),
                    data=df.MT.total)
```

### Diagnostics
```{r}
sim.MT.total <- simulateResiduals(fittedModel = MT.total.glmm.zi.complex, n=1000)
plot(sim.MT.total, quantreg = TRUE)
```


### Significances

**receive the variance of each factor of the model**
```{r}
var.glmm.MT.total <- get_variance(MT.total.glmm.zi.complex, component = "all")
print(var.glmm.MT.total)
```


**Chi-square test**

```{r}
anova.glmm.MT.total <- glmmTMB:::Anova.III.glmmTMB(MT.total.glmm.zi.complex,
                                                  component = "cond")
anova.glmm.MT.total.zi <- glmmTMB:::Anova.III.glmmTMB(MT.total.glmm.zi.complex,
                                                      component = "zi")
print(anova.glmm.MT.total)
print(anova.glmm.MT.total.zi)
```

**Emmean estimates of significant interaction in zi-component of model**

```{r}
MT.count.zi.emmeans <- as.data.frame(emmeans(MT.total.glmm.zi.complex, pairwise~Chemotype * Plottype, 
                                          type = "response",
                                          component = "zi")) %>%
  dplyr::select(c("Chemotype", "Plottype", "response", "SE")) %>%
  filter(!(Chemotype == "." & Plottype == ".")) %>%
  mutate(Chemotype = factor(Chemotype, levels = c("Keto",
                                                  "BThu",
                                                  "ABThu",
                                                  "Aacet",
                                                  "Myrox")),
         Plottype = factor(Plottype, levels = c("Homogenous",
                                                "Heterogenous")),
         errorbarmin = response - SE,
         errorbarmax = response + SE)
```

**Plot the emmean estimates**

```{r}
gg.MT.count.lsmean <- ggplot(aes(x = Chemotype, 
                                 y = response, 
                                 fill = Chemotype,
                                 alpha = Plottype),
             data= MT.count.zi.emmeans) +
  geom_bar(stat="identity", position="dodge") +
  geom_errorbar(aes(ymin=response-SE, ymax=response+SE),
                width=.3, position=position_dodge(.9)) +
  ylim(0,6) +
  scale_x_discrete("\nChemotype") +
  scale_fill_manual(values = colours_chemotype, guide="none") +
  scale_alpha_discrete(name = "Plot-type",
                       range = c(1, 0.4)) +
  scale_y_continuous(expression(paste("Probability total presence  ",italic("M. tanacetaria"))),
                     limits = c(0, 1),
                     breaks = seq(0, 1, by = 0.2) ) +
  guides(alpha = guide_legend(override.aes = list(fill="grey", alpha = c(1, 0.5),
                                                  linetype = 0))) +
  theme_classic() +
  theme(text = element_text(size=12),
        axis.title.y = element_text(vjust = +3))

show(gg.MT.count.lsmean)
```

## Alate
```{r}
df.MT.alate <- data %>%
  filter(Week > 20,
         Week < 27)
```

### Binomial model
```{r}
glmm.MT.alate.presence <- glmmTMB(M_tanacetaria_alate_presence ~ Chemotype*Plottype + Week 
                                  + (1|Block) 
                                  + (1|Plot) 
                                  + (1|Clone_ID) 
                                  + (1|Motherplant) 
                                  + (1|Counted_by) 
                                  + (1|Plant),
                    data=df.MT.alate,
                    family="binomial")
```

### Diagnostics
```{r}
sim.MT.alate <- simulateResiduals(fittedModel = glmm.MT.alate.presence, n=1000)
plot(sim.MT.alate, quantreg = TRUE)
```


### Significances
**Get the variance explained by each factor**
```{r}
var.glmm.MT.alate.presence <- get_variance(glmm.MT.alate.presence, component = "all")
var.glmm.MT.alate.presence
```


**Chi-square test**
```{r}
anova.glmm.MT.alate.presence <- Anova(glmm.MT.alate.presence, type="III",test="Chisq")
print(anova.glmm.MT.alate.presence, digits = 3)
```
# *M. fuscoviride* 

## Total count
```{r}
data.MF.total <- data %>%
  filter(Week > 21,
         Week < 28)
```

### The zero-inflation model
```{r}
glmm.MF.zi.total <- glmmTMB(M_fuscoviride_total ~ Chemotype * Plottype + Week +
                                      Ant_total_presence + Antnest_presence + Ant_MF_presence
                    #+ (1|Block) # Variance explained too low, causes singularity problems
                    + (1|Plot)
                    + (1|Clone_ID)  
                    #+ (1|Motherplant) # Variance explained too low, causes singularity problems
                    + (1|Plant),
                    family = "nbinom2",
                    zi=~Chemotype * Plottype + Week +
                      Ant_total_presence + Antnest_presence + Ant_MF_presence
                    #+ (1|Block) # Variance explained too low, causes singularity problems
                    + (1|Plot)
                    + (1|Clone_ID) 
                    #+ (1|Motherplant) # Variance explained too low, causes singularity problems
                    + (1|Plant),
                    data=data.MF.total)
```

### Diagnostics
```{r}
sim.glmm.MF.zi.total <- simulateResiduals(glmm.MF.zi.total, n = 1000)
plot(sim.glmm.MF.zi.total, quantreg=T)
```


### Significances
**receive the variance of each factor of the model**
```{r}
var.glmm.MF.zi.total <- get_variance(glmm.MF.zi.total)

print(var.glmm.MF.zi.total)

```

**Chi-square test**
```{r}
anova.zi.glmm.MF.zi.total <- glmmTMB:::Anova.glmmTMB(glmm.MF.zi.total,
                                                     type = "III",
                                                     component = "zi",
                                                     test.statistic = "Chisq")

anova.cond.glmm.MF.zi.total <- glmmTMB:::Anova.glmmTMB(glmm.MF.zi.total,
                                                       type = "III",
                                                       component = "cond",
                                                       test.statistic = "Chisq")

print(anova.cond.glmm.MF.zi.total, digits = 3)
print(anova.zi.glmm.MF.zi.total, digits = 3)
```

## Alate

```{r}
data.MF.alate <- data %>%
  filter(Week > 19,
         Week < 27)
```

### Binomial model
```{r}
glmm.MF.alate.presence <- glmmTMB(M_fuscoviride_alate_presence ~ Chemotype * Plottype + Week +
                                    Ant_total_presence + Antnest_presence + Ant_MF_presence
                                 + (1|Block) 
                                 + (1|Plot) 
                                 + (1|Clone_ID) 
                                 + (1|Motherplant)
                                 + (1|Counted_by) 
                                 + (1|Plant),
                    data=data.MF.alate, 
                    family="binomial")
```


### Diagnostics

```{r}
sim.MF.alate.glmm <- simulateResiduals(glmm.MF.alate.presence, n = 1000)
plot(sim.MF.alate.glmm, quantreg=T)
```


### Significances

**receive the variance of each factor of the model**
```{r}
var.MF.alate.glmm <- get_variance(glmm.MF.alate.presence)
var.MF.alate.glmm
```

**Chi-square test**
```{r}
anova.MF.alate.glmm <- Anova(glmm.MF.alate.presence, type="III", test="Chisq")
print(anova.MF.alate.glmm, digits = 3)
```

# Combining both significant emmeans plots

When knitting the document, it seems the patchwork code hands out an error. When
run individually in RStudio however it works. Therefore, the code is included as code
but without output to allow for knitting this document as html.
If you want to see the output, please check Figure 2 in the plot-folder.
Alternatively you can run the script by yourself individually by directly using the
03_GLMMs_wholeseason.Rmd file

```{r Combining both significant emmeans plots, eval=F, echo=T}
barplot.lsmeans.ut.total.merge <- barplot.lsmeans.ut.total +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        text = element_text(size = 11),
        legend.text = element_text(size = 8)) +
  coord_cartesian(clip = "off") +
    annotate(geom = "text", x = 0, y = 150,
            label = "A", hjust = 3.4, vjust = -1.3, size = 4, fontface = "bold")

gg.MT.count.lsmean.merge <- gg.MT.count.lsmean +
  theme(text = element_text(size = 11),
        legend.text = element_text(size = 8)) +
  coord_cartesian(clip = "off") +
  annotate(geom = "text", x = 0, y = 1,
            label = "B", hjust = 3.4, vjust = -1.3, size = 4, fontface = "bold")

LSmeans.combined.MT.UT <- barplot.lsmeans.ut.total.merge / gg.MT.count.lsmean.merge  +
                                plot_layout(guides = "collect") & theme(legend.position = "top")

show(LSmeans.combined.MT.UT)

ggsave("Plots/Figure_2.pdf", LSmeans.combined.MT.UT,
       dpi=600,
       device = "pdf",
       width = 8.5, 
       height = 16.7,
       units = "cm")

ggsave("Plots/Figure_2.jpg", LSmeans.combined.MT.UT,
       dpi=600,
       width = 8.5, 
       height = 16.7,
       units = "cm")
```

