---
title: "Lasso regression of terpenoids on Aphid abundance significantly affected by Shannon-diversity"
author: "Dominik Ziaja"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    number_section: true
    df_print: paged
    theme: cerulean
---

Notebook to analyse the importance of certain terpenoids in the abundance of the different aphid specialist species

```{r}
library(glmnet)
library(car) # qqPlot (distribution = "pois" used for distribution checking for LASSO Regression)
library(tidyverse)
library(fitdistrplus)
```
# Load in necessities

## set seed

```{r}
set.seed(100)
```



## Functions used
```{r}
df.lasso.tidiying <- function(df, Kept.aphidcolumn){
  data.lasso <- df %>%
  dplyr::select(c(Kept.aphidcolumn, "cadinene_g":"santolina_triene"))
  
  return(data.lasso)}
```

## Packages

```{r}
df.aphid <- as_tibble(read.csv("Data/Counteddata_ComGar_2021_tidied.csv"))
df.terpenes <- as_tibble(read.csv("Data/Phenotypingweek_terpenedata_tidied.csv"))
df.morphology <- as_tibble(read.csv("Data/Phenotypingweek_Morphologydata.csv"))
```

# Data tidying and wrangling

## aphid data

```{r}
df.aphid <- df.aphid %>%
  filter(Week == 25) %>%
    dplyr::select(c("Block":"Plottype",
           contains(c("M_fuscoviride", "U_tanaceti", "M_tanacetaria",
                      "Antnest_presence", "Ant_MF_presence", "Ant_total_presence")),
           "Counted_by"),
           -"Plant")
```

## terpene data

```{r}
df.terpenes <- df.terpenes %>%
  filter(!(Block == 1 & Plot == 1)) %>%
  dplyr::select(-c("Comp_ID", "Norm_weight", "Peak_area")) %>%
  pivot_wider(names_from = Comp_name, values_from = Percent)
```

## morphology data

```{r}
df.morphology <- df.morphology %>%
  filter(!(Block == 1 & Plot == 1))
  
```

## Construct dataframe with metainfo

```{r}
metainformation <- df.morphology %>%
  dplyr::select("Block":"Chemotype") %>%
  inner_join(., df.aphid[,c("Plant_standardized", "Plot", "Position", "Plottype")], by=c("Plant" = "Plant_standardized")) %>%
  dplyr::rename("Plot_unique" = "Plot.y",
         "Position_unique" = "Position.y",
         "Plot_redundant" = "Plot.x",
         "Position_redundant" = "Position.x")

# set up a dataframe containing all the metainformation for the different plots
metainformation.plotlevel <- metainformation %>%
  filter(duplicated(Plot_unique) == FALSE) %>%
  dplyr::select(-c("Position_unique", "Position_redundant")) %>%
  mutate(Chemotype = case_when(Plottype == "Mixed" ~ "Mixed",
                               Plottype == "Mono" ~ Chemotype))
  
    
```

## Data check
Check whether in all dataframes the plant IDs are equal
This allows us to later on merge the data frames based on the sample_IDs 
```{r}
all(df.aphid$Plant_standardized %in% df.morphology$Plant) & all(df.terpenes$Plant %in% df.aphid$Plant_standardized)
```

## Construct merged aphid and terpenoid dataframe

```{r}
df.aphid.terpenes <- df.aphid %>%
  inner_join(df.terpenes, by = c("Plant_standardized" = "Plant")) %>%
  dplyr::select(c("Plot.x",
           contains(c("alate_presence", "total")), 
           "cadinene_g":"yomogi_alcohol"),
         -(contains("Ant")),
         "chrysanthenol",
         "chrysanthenone",
         "chrysanthenyl_acetate_E",
         "chrysanthenyl_acetate_Z",
         "santolina_triene") %>%
  rename("Plot" = "Plot.x")

df.aphid.terpenes.lasso <- df.aphid.terpenes %>%
  dplyr::select(-("Plot"))
```

# Lasso-regression

## *M. tanacetaria*

### Distribution estimate

```{r}
pois.estimates.mtanacetariatotal <- fitdist(df.aphid.terpenes$M_tanacetaria_total, "pois")
plot(pois.estimates.mtanacetariatotal)
qqPlot(df.aphid.terpenes$M_tanacetaria_total, distribution = "pois", lambda = pois.estimates.mtanacetariatotal$estimate)
```

**Seems like especially the counts higher than ~110 deviate heavily from the poisson distribution**


### Lasso calculation

```{r}

data.lasso <- df.lasso.tidiying(df.aphid.terpenes, "M_tanacetaria_total")

y <- data.lasso %>%
  filter(M_tanacetaria_total < 150) %>%
  dplyr::select("M_tanacetaria_total") %>%
  as.matrix()

x <- data.lasso %>%
  filter(M_tanacetaria_total < 150) %>%
  dplyr::select("cadinene_g":"santolina_triene") %>%
  as.matrix()


#lambda.min <- average_lasso_crossvalidation(y, x, family = "poisson")

lasso.fit <- cv.glmnet(y=y, x=x, 
                       #lambda = lambda.min, 
                       family = "poisson")

lasso.fit.matrix <- as.matrix(coef(lasso.fit, s = "lambda.1se"))

print(subset(lasso.fit.matrix, lasso.fit.matrix > 0))
```

**Only an intercept estimate is given -> No linear combination available**

## *U. tanaceti*

### Dataframe preparation

**First the data on the plot level needs to be calculated**

```{r}
df.terpenes.plot <- df.aphid.terpenes %>%
  dplyr::select(c("Plot", "cadinene_g":"santolina_triene")) %>%
  group_by(Plot) %>%
  summarise_all(mean)

df.aphids.plot <- df.aphid.terpenes %>%
  dplyr::select(c("Plot", "U_tanaceti_total")) %>%
  group_by(Plot) %>%
  summarise_all(sum)

df.aphid.terpenes.plot <- df.aphids.plot %>%
  inner_join(., df.terpenes.plot, by = c("Plot" = "Plot"))
```


### Distribution estimate

```{r}
pois.estimates.utanacetitotalplot <- fitdist(df.aphid.terpenes.plot$U_tanaceti_total, "pois")
plot(pois.estimates.utanacetitotalplot)
qqPlot(df.aphid.terpenes.plot$U_tanaceti_total, distribution = "pois", lambda = pois.estimates.utanacetitotalplot$estimate)
```

### Lasso calculation
 
```{r}
data.lasso.plot <- df.lasso.tidiying(df.aphid.terpenes.plot, "U_tanaceti_total")

y <- data.lasso.plot %>%
  dplyr::select("U_tanaceti_total") %>%
  filter(U_tanaceti_total < 4000) %>%
  as.matrix()

x <- data.lasso.plot %>%
  filter(U_tanaceti_total < 4000) %>%
  dplyr::select("cadinene_g":"santolina_triene") %>%
  as.matrix()


lasso.fit <- cv.glmnet(y=y, x=x, 
                       family = "poisson")

lasso.fit.matrix <- as.matrix(coef(lasso.fit, s = "lambda.1se"))

print(subset(lasso.fit.matrix, lasso.fit.matrix > 0))
```

**Only an intercept estimate is given -> No linear combination available**